---
title: "ShoshanaBuck_challenge 1"
author: "Shoshana Buck"
desription: "Reading in data and creating a post"
date: "08/15/2022"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_1
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to

1)  read in a dataset, and

2)  describe the dataset using both words and any supporting information (e.g., tables, etc)

## Read in the Data

Read in one (or more) of the following data sets, using the correct R package and command.

-   railroad_2012_clean_county.csv ⭐
-   birds.csv ⭐⭐
-   FAOstat\*.csv ⭐⭐
-   wild_bird_data.xlsx ⭐⭐⭐
-   StateCounty2012.xlsx ⭐⭐⭐⭐

Find the `_data` folder, located inside the `posts` folder. Then you can read in the data, using either one of the `readr` standard tidy read commands, or a specialized package such as `readxl`.

```{r}
railroad = read.csv('C:\\Users\\user\\Desktop\\dacss601_august2022\\posts\\_data\\railroad_2012_clean_county.csv', header = TRUE)

```

Add any comments or documentation as needed. More challenging data sets may require additional code chunks and documentation.

## Describe the data

Using a combination of words and results of R commands, can you provide a high level description of the data? Describe as efficiently as possible where/how the data was (likely) gathered, indicate the cases and variables (both the interpretation and any details you deem useful to the reader to fully understand your chosen data).

```{r}
#| label: summary

#dim dimentions of the dataset, number of rows and columns
dim(railroad)

# str, number of observations, variables (columns) and class of each columns variable, along with few observations

str(railroad)

#summary, summary of railroad dataset of individual column variables; with length, class & mode for character class columns variables and quartile distribution for numeric class column variable along with min-max values
summary(railroad)

# psych describe
# displays: number of observations, mean, standard distribution (sd), median, trimmed, mad, min, max values for the column variables

psych::describe(railroad)

#Hmisc describe
# displays number of observations,  number of missing value in columns, number of distinct values in columns, Mean Gmd, lowest, highest values etc.

Hmisc::describe((railroad))

```
