---
title: "Challenge 1"
author: "Quinn He"
desription: "Reading in data and creating a post"
date: "08/15/2022"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_1
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to

1)  read in a dataset, and

2)  describe the dataset using both words and any supporting information (e.g., tables, etc)

## Read in the Data

Read in one (or more) of the following data sets, using the correct R package and command.

-   railroad_2012_clean_county.csv ⭐
-   birds.csv ⭐⭐
-   FAOstat\*.csv ⭐⭐
-   wild_bird_data.xlsx ⭐⭐⭐
-   StateCounty2012.xlsx ⭐⭐⭐⭐

Find the `_data` folder, located inside the `posts` folder. Then you can read in the data, using either one of the `readr` standard tidy read commands, or a specialized package such as `readxl`.

```{r}

birds <- read_csv("../posts/_data/birds.csv", show_col_types = FALSE)

view(birds)

```

Add any comments or documentation as needed. More challenging data sets may require additional code chunks and documentation.

## Describe the data

Using a combination of words and results of R commands, can you provide a high level description of the data? Describe as efficiently as possible where/how the data was (likely) gathered, indicate the cases and variables (both the interpretation and any details you deem useful to the reader to fully understand your chosen data).

```{r}
#| label: The birds dataset contains a wide range of range of entries. With the function below we can see all the column names listed. A few are hard to figure out what exactly the represent and just how important they are. 

colnames(birds)

```

```{r}
#| label: It appears the dataset was taken from a farm organization. The data is definitely a little messy, but makes sense on the data entry side. Each country has decsending rows of chickens, ducks, and fowls from 1961 to 2018. This is mostly a bit redundant. This whole dataset keeps track of the value of these three types of birds in a 60 year window. There is also a possibility this dataset came from a larger set with other types of animals becuase the "Domain" column lists 'Livestock' throughout the entire dataset. 




```
